---
title: "The Relationship Between Life Expectancy and the Gini Coefficient"
author:
  - "Vishwamber Reddy"
  - "Caoyu Shao"
  - "Hongyu Yin"
toc: true
format: html
---

# **Executive summary**

Studying the relationship between life expectancy and the Gini coefficient is essentially exploring the connection between economic inequality and health inequality.

In this era that emphasizes economic development, people's physical health should be equally prioritized.

By comparing the Gini coefficients and life expectancies across different economies and regions, we can gain a more intuitive and comprehensive understanding of a particular economy or region's development level.

Furthermore, researching the relationship between these inequalities can provide new directions for economic development goals.

# **Introduction**

## **Research Question**

In this study, we will explore whether reducing economic inequality contributes to mitigating disparities in life expectancy.

The Gini coefficient is a key indicator for measuring economic inequality within a country or region.Therefore, by examining the relationship between the Gini coefficient and life expectancy, we aim to assess the importance of balanced economic development.

## **Dataset introduction**

We utilize **life expectancy** data from **Our World** in Data and **Gini coefficient** data from **the World Bank**, spanning 1990 to 2020 and covering multiple countries and regions.

-   The body of the report should contain the following sections:

    **Methodology**: How we processed the data.

    **Results**: What conclusions we reached.

    **Discussion, conclusion and recommendations**: Our Intended Message.

# **Methodology**

```{r}
# 1. Install & load the WDI package
if (!require(WDI)) install.packages("WDI")
library(WDI)

# 2. Define WDI indicator codes
indicators <- c(
  GDPpc       = "NY.GDP.PCAP.CD",     
  EduTertiary = "SE.TER.ENRR",       
  HealthExp   = "SH.XPD.CHEX.GD.ZS",  
  UrbanShare  = "SP.URB.TOTL.IN.ZS",  
  SmokePrev   = "SH.PRV.SMOK",        
   Gini    = "SI.POV.GINI",    
  LifeExp = "SP.DYN.LE00.IN"  )

variables <- WDI(
  country   = "all",
  indicator = indicators,
  start     = 2005,
  end       = 2025,
  extra     = FALSE     # adds iso codes, region, income groups, etc.
)

head(variables)
```

```{r}
#pearson test, which is aimed to test the correlations betweent any two variables
install.packages("Hmisc")
library(Hmisc)

#    LifeExp, Gini, GDPpc, EduTertiary, HealthExp, UrbanShare, SmokePrev
vars <- c("LifeExp", "Gini", "GDPpc", "EduTertiary",
          "HealthExp", "UrbanShare", "SmokePrev")


mat <- as.matrix(variables[, vars])

#calculate the P value
res <- rcorr(mat, type = "pearson")


# res$r is the correlation matrix
print(res$r)

# res$P is the corresponding matrix
print(res$P)

```

# **Results**

Maximum 200 words. Should include either a figure or a table.

# **Discussion, conclusion and recommendations**
According to our discussion, we get a conclusion that there is a negative 
relationship between gini-coefficient and life-expectation. 
We get four explanations for that.

First, due to resource inequality, there will be more poor people and 
less rich people if there is a high gini coefficient. Rich people could afford
higher-quality care, preventive screenings, and insurance, while 
poorer people may delay or forgo treatment due to cost. And besides Affluent 
households can buy healthier food and live in cleaner, less crowded 
environments; low-income families face greater exposure to pollution, 
overcrowding, and food insecurity.

Second, due to Psychosocial Stress and Social Capital, large income gaps fuel 
social comparison and status anxiety, triggering persistent stress responses 
that harm cardiovascular, immune, and endocrine systems. What's more, 
High-inequality societies often exhibit lower trust and mutual support, 
reducing the informal networks that share health information and 
provide neighborhood‐level care.

Third, due to underinvestment in Public Health and Welfare.When tax, welfare,
and public services disproportionately benefit the well-off, funding for 
universal healthcare, elder care, and child health programs suffers—limiting 
gains in overall longevity.Besides, Education strongly influences health 
literacy and behaviors; in unequal societies, lower-educated groups 
lack knowledge about disease prevention and healthy lifestyles, increasing
chronic‐disease risks.

Lastly,due to cumulative and Intergenerational Effect, Childhood 
poverty—malnutrition, exposure to toxins, or family financial stress—can 
stunt physical and cognitive development, with effects that persist into 
adulthood.Besides, in highly unequal settings, children from poor families 
have fewer opportunities to escape poverty, perpetuating health 
disadvantages across generations.

# **Reference section**

De Vogli, R., Mistry, R., Gnesotto, R., & Cornia, G. A. (2005). Has
    the relation between income inequality and life expectancy
    disappeared? Evidence from Italy and top industrialised countries.
    Journal of Epidemiology & Community Health, 59(2), 158–162.
    https://doi.org/10.1136/jech.2004.020651 ResearchGate

-   Preston, S. H. (1975). The changing relation between mortality and
    level of economic development. Population Studies, 29(2), 231–248.
    JSTOR

-   Hazan, M. (2012). Life expectancy and schooling: new insights from
    cross‐country data. Journal of Population Economics, 25, 1237–1248.
    https://doi.org/10.1007/s00148-011-0392-6 SpringerLink

-   Jaba, E., Bălan, C. B., & Robu, I.-B. (2014). The relationship
    between life expectancy at birth and health expenditures estimated
    by a cross‐country and time‐series analysis. Procedia Economics and
    Finance, 15, 108–114. DergiPark

-   Tripathi, S., & Maiti, M. (2023). Does urbanization improve health
    outcomes: a cross country level analysis. Asia‐Pacific Journal of
    Regional Science, 7, 277–316.
    https://doi.org/10.1007/s41685-022-00268-1 SpringerLink

-   GBD 2015 Tobacco Collaborators. (2017). Smoking prevalence and
    attributable disease burden in 195 countries and territories,
    1990–2015: a systematic analysis from the Global Burden of Disease
    Study 2015. The Lancet, 389(10082), 1885–1906.
    https://doi.org/10.1016/S0140-6736(17)30819-X thelancet.com


